Deep mutational scanning of hemagglutinin helps distinguish the evolutionary fates of human H3N2 influenza virus lineages

### 2018-01-30 paper/manuscript.tex
Minor revisions

\affil[a]{Basic Sciences Division}
\affil[b]{Vaccine and Infectious Diseases Division}
\affil[c]{and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA}
\affil[d]{Department of Genome Sciences}
\affil[e]{Medical Scientist Training Program}
\affil[f]{and Molecular and Cellular Biology Program, University of Washington, Seattle, WA 98195, USA}

@@

\subsection*{The experimental measurements can help discriminate between successful and unsuccessful influenza virus lineages}
A major goal in the study of rapidly evolving viruses such as influenza is to forecast evolutionary trajectories~\cite{lassig2017predicting,morris2017predictive}.
For instance, forecasting which viral lineages will dominate the upcoming influenza season is an integral part of vaccine-strain selection~\cite{neher2015nextflu,lassig2017predicting,morris2017predictive}. 
Evolutionary forecasts must ultimately distinguish between successful and unsuccessful viral lineages, which in the case of human influenza virus means distinguishing between the trunk and side branches of the phylogenetic tree (Figure~\ref{fig:H3N2_phylogeny}).

To investigate whether our experiments can aid in distinguishing trunk and side branch lineages, we calculated the experimentally measured effects of all mutations in a maximum-likelihood reconstruction of the phylogeny in Figure~\ref{fig:H3N2_phylogeny}.
The mutations that occurred on the trunk of the tree were consistently more beneficial for viral growth according to our experimental measurements (Figure~\ref{fig:trunkvssidebranch}A).
This difference in the effects of mutations between the trunk and side branch lineages was statistically significant when taken across the entire phylogeny (Figure~\ref{fig:trunkvssidebranch}B).
Some influenza sequences are derived from egg- or cell-passaged isolates that contain lab-adaptation mutations~\citep{mcwhite2016sequence}.
These mutations mostly occur on the terminal side branches that lead to tip nodes on the phylogenetic tree.
We therefore re-calculated the statistics separating side branches to tip and internal nodes, and again found that the difference between the trunk and both sets of side branches was statistically significant (Figure~\ref{fig:trunkvssidebranch}B).
Therefore, strains with mutations that we experimentally measure to be favorable for viral growth tend to do better in nature than strains with mutations that we measure to be less favorable.

@@

\subsection*{Inference of human H3N2 phylogenetic tree}
We applied Nextstrain's augur pipeline~\cite{Hadfield224048} (https://github.com/nextstrain/augur; commit 006896d) to publicly available H3N2 HA sequences from GISAID, sampling six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
We aligned the resulting 2,139 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.
Ancestral state reconstruction and branch length timing were performed with treetime~\cite{Sagulenko2018}.
We pruned seven strains with premature stop codons in their amino acid translations that represented likely sequencing errors, producing a final tree with 2,132 strains.

To parse trunk mutations from side branch mutations, we defined a set of recent nodes sampled on or after January 1, 2017, and traced these nodes back to their most recent common ancestor.
We defined all branches ancestral to this most recent common ancestor as the trunk.
All nodes that do not have descendants isolated after 2017 were defined as side branch nodes.
Nodes descended from the most recent common ancestor were defined as unresolved, as it is still unclear whether these nodes will ultimately be on the trunk or side branches.
The phylogenetic tree is available as a JSON file on \comment{link to repository}.

@@

\acknow{We thank Sarah Hilton, Hugh Haddox, and Sidney Bell for helpful discussions about data analysis. 
We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing. 
This work was supported by grant R01AI127897 from the NIAID of the NIH. 
JML was supported in part by the Center for Inference and Dynamics of Infectious Diseases (CIDID), which is funded by grant U54GM111274 from the NIGMS of the NIH.
The research of JDB is supported in part by a Faculty Scholar grant from the Howard Hughes Medical Institute and the Simons Foundation, and a Burroughs Wellcome Young Investigator in the Pathogenesis of Infectious Diseases grant.
\comment{Add any additional relevant grants for Trevor and John}
}

>>>>

\affil[a]{Basic Sciences Division}
\affil[b]{Vaccine and Infectious Disease Division}
\affil[c]{and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA}
\affil[d]{Department of Genome Sciences}
\affil[e]{Medical Scientist Training Program}
\affil[f]{and Molecular and Cellular Biology Program, University of Washington, Seattle, WA 98195, USA}

@@

\subsection*{The experimental measurements can help discriminate between successful and unsuccessful influenza virus lineages}
A major goal in the study of rapidly evolving viruses such as influenza is to forecast evolutionary trajectories~\cite{lassig2017predicting,morris2017predictive}.
For instance, forecasting which viral lineages will dominate the upcoming influenza season is an integral part of vaccine-strain selection~\cite{neher2015nextflu,lassig2017predicting,morris2017predictive}.
Evolutionary forecasts may target fixation and loss of alleles in the viral population, which means distinguishing between the trunk and side branches of the phylogenetic tree (Figure~\ref{fig:H3N2_phylogeny}).

To investigate whether our experiments can aid in distinguishing trunk and side branch lineages, we calculated the experimentally measured effects of all mutations in a maximum-likelihood reconstruction of the H3N2 phylogeny (Figure~\ref{fig:H3N2_phylogeny}).
The mutations that occurred on the trunk of the tree were consistently more beneficial for viral growth according to our experimental measurements (Figure~\ref{fig:trunkvssidebranch}A).
This difference in the effects of mutations between the trunk and side branch lineages was statistically significant when taken across the entire phylogeny (Figure~\ref{fig:trunkvssidebranch}B).
Some influenza sequences are derived from egg- or cell-passaged isolates that contain lab-adaptation mutations~\citep{mcwhite2016sequence}.
These mutations mostly occur on the terminal side branches that lead to tip nodes on the phylogenetic tree.
We therefore re-calculated the statistics separating side branches to tip and internal nodes, and again found that the difference between the trunk and both sets of side branches was statistically significant (Figure~\ref{fig:trunkvssidebranch}B).
Therefore, strains with mutations that we experimentally measure to be favorable for viral growth tend to do better in nature than strains with mutations that we measure to be less favorable.

@@

\subsection*{Inference of human H3N2 phylogenetic tree}
\comment{John to add methods here. We downloaded X sequences from the Influenza Virus Resource ?.... etc. subsampled how? inferred the tree, ancestral state reconstruction, visualized the tree...}
To parse trunk mutations from side branch mutations, we defined a set of recent nodes sampled on or after Jan-1-2017, and traced these nodes back to their most recent common ancestor. 
We defined all branches ancestral to this most recent common ancestor as the trunk.
All nodes that do not have descendants isolated after 2017 were defined as side branch nodes.
Nodes descended from the most recent common ancestor were defined as unresolved, as it is still unclear whether these nodes will ultimately be on the trunk or side branches.
The phylogenetic tree is available as a JSON file on \comment{link to repository}.

@@

\acknow{We thank Sarah Hilton, Hugh Haddox, and Sidney Bell for helpful discussions about data analysis.
We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
This work was supported by grant R01 AI127893 from the NIAID of the NIH to JDB and TB and grant U19 AI117891 to TB.
JML was supported in part by the Center for Inference and Dynamics of Infectious Diseases (CIDID), which is funded by grant U54GM111274 from the NIGMS of the NIH.
The research of JDB is supported in part by a Faculty Scholar grant from the Howard Hughes Medical Institute and the Simons Foundation, and a Burroughs Wellcome Young Investigator in the Pathogenesis of Infectious Diseases grant.
TB is a Pew Biomedical Scholar and is supported by NIH R35 GM119774.
}
